---
title: "Homework Assignment 4: Tidy Data and Joins "
author: "Hannah McSorley"
date: "Due 2019-10-08"
output: 
  html_document:
    keep_md: true
    theme: cerulean
---

## Assignment Objective

The focus of this assignment is to practice problem solving in R using two key components of data wrangling: data aggregation and reshaping. In addition, data visualization with ggplo2 will be included. 
The data used in this assignment includes 'gapminder' (for Exercises 1 & 2), and two wedding-planning data sets (for Exercise 3).

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.align = "center")
```

```{r packages, warning = FALSE, message = FALSE}
library(tidyverse)
library(gapminder)
library(knitr)  # for tidy tables
library(DT)     # for datatable in html
```

# Exercise 1: Univariate Data Reshaping 

__Option 2__

Three Goals:

- 1. Reshape data into wider format
- 2. Produce a plot of the wide data
- 3. Re-lengthening the wider data

This code creates two wide tibbles, with years/continents as either rows or columns.

```{r E1 wider}
# mutate and reshape data into wider format
## use tidyr::pivot_wider()
# A: rows = year, columns = continents
A_wide <- gapminder %>% 
  dplyr::arrange(continent, year) %>% 
  group_by(continent, year) %>% 
  summarize(med_lifeExp = median(lifeExp)) %>% 
  tidyr::pivot_wider(id_cols = year,
                         names_from = continent,
                         values_from = med_lifeExp)
# table
knitr::kable(A_wide, digits = 0, 
      caption = "Table 1A-Wide: continental median life expectancy")

# B: rows = continents, columns = years
B_wide <- gapminder %>% 
  dplyr::arrange(continent, year) %>% 
  group_by(continent, year) %>% 
  summarize(med_lifeExp = median(lifeExp)) %>% 
  tidyr::pivot_wider(id_cols = continent,
                         names_from = year,
                         values_from = med_lifeExp) 
# table
knitr::kable(B_wide, digits = 0, 
      caption = "Table 1B-Wide: median life expectancy over time by continent")
```

Visually, it might be more intuitive to read Table 1B-Wide, with years progressing from left to right. However, it seems difficult to plot from that tibble shape, especially because the variables are numbers (years). 

```{r E1 wide plot}
# Produce a plot of (some of) the wide data
# B (easy to read)
B_wide %>% ggplot(aes(x = 1952, y = 2007)) + 
  geom_point(colour = "magenta2") +
  theme_light()
# the data from Table 1-B is difficult to plot (at least for me)
# the variables are numbers (years), and are treated as discrete numbers in ggplot

# try plotting the data from Table 1A-Wide
# A
A_wide %>% ggplot(aes(x = year, y = Africa)) + 
  geom_point(colour = "steelblue4") +
  theme_light()
# it's easier to plot a contient's median life expectancy over time from Table 1-A
# trickier to plot multiple contients over time though...that's easier from long format

# easy to plot one continent against another
A_wide %>% ggplot(aes(x = Americas, y = Asia)) + 
  geom_point(colour = "tomato") +
  theme_light()
```

It seems that the data shape which is easiest to read, or most intuitive to present, is the most difficult to plot in R. Whereas longer tibbles are easier for R to read and plot. 

```{r E1 longer}
# Re-lengthening the wider data
## use tidyr::pivot_longer()
#A
A_long <- A_wide %>% 
  tidyr::pivot_longer(cols = Africa:Oceania,
                      names_to = "continent",
                      values_to = "median_lifeExp") %>% 
  arrange(continent)
# table
DT::datatable(A_long, caption = "Table 1A-Long: continental median life expectancy")
  
# check that you produced what you wanted:
gapminder %>% 
  dplyr::arrange(continent, year) %>% 
  group_by(continent, year) %>% 
  summarize(med_lifeExp = median(lifeExp))
# Yes!

# Re-lengthen the data from Table 1-B also
B_long <- B_wide %>% 
  tidyr::pivot_longer(cols = (2:13),
                      names_to = "year",
                      values_to = "median_lifeExp") %>% 
  arrange(continent)
# table
DT::datatable(B_long, caption = "Table 1B-Long: continental median life expectancy")

```

# Exercise 2: Multivariate Data Reshaping 

Two Goals:

- 1. Reshape data into wider format
- 2. Re-lengthening the wider data

__Option 1__ 
_Specifically_

- 1. Make a tibble with one row per year, and columns for life expectancy and GDP per capita (or two other numeric variables) for two or more countries.
- 2. Re-lengthen the data


```{r E2 wider}
# filter for two countries' lifeExp & gdpPercap
# pivot_wider()
# then pivot back (pivot_longer())
E2_wide <- gapminder %>% 
  dplyr::arrange(country, year) %>% 
  select(-pop, -continent) %>%
  filter(country == "Guinea" | 
           country == "Slovenia") %>%
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = c(lifeExp, gdpPercap))
knitr::kable(E2_wide, digits = c(0, 0, 0, 2, 2), 
             caption = "Wide format: multivariate countries' GDP and life expectancy")

```

Take the wider tibble and pivot it back to long format. 
```{r E2 longer}
# filter for two countries' lifeExp & gdpPercap
# pivot_wider()
# then pivot back (pivot_longer())
E2_longer <- E2_wide %>% 
   pivot_longer(cols = -year,
              names_to = c(".value", "Country"),
              names_sep = "_",
              values_to = c("lifeExp", "gdpPercap")) %>% 
   arrange(Country)
knitr::kable(E2_longer, digits = c(0, 0, 0, 2, 2), 
             caption = "Wide format: multivariate countries' GDP and life expectancy")
```


# Exercise 3: Table Joins

Read in the made-up wedding guestlist and email addresses:

```{r E3 data}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")

# check the top of each data set:
head(guest)
head(email)
```

## 3.1

For each guest in the guestlist (guest tibble), add a column for email address, which can be found in the email tibble.

___Note:___
Before joining these tibbles, I separated names that were listed (probably by wedding party) in the tibble 'email'. I found how to use 'strsplit()' with 'unlist()' to accomplish the unlisting on [StackOverflow](https://stackoverflow.com/questions/15347282/split-delimited-strings-in-a-column-and-insert-as-new-rows).


```{r E3.1}
# mutating join
# add data from 'email' to 'guest'
guest %>% 
  left_join(email %>% 
              mutate(guest = strsplit(as.character(guest), ", ")) %>% 
              unnest(guest), # split the lists of guests from tibble 'email' 
            by = c("name" = "guest")) %>%   # the two tibbles have different key names
  DT::datatable(caption = "guestlist with emails")
```


## 3.2 

Who do we have emails for, yet are not on the guestlist?

I used _anti_join()_, a filtering join that drops observations from the primary tibble (x = 'email') which don't have a match in the joining/reference tibble (y = 'guest').

Like E3.1, this code splits the string of names in the email data set into their own row before joining tibbles.

```{r E3.2}
# filtering join
# find 'email' guests who are not on the 'guests' list
# anti_join(x, y) drops all observations in x that have a match in y
email %>% 
  mutate(guest = strsplit(as.character(guest), ", ")) %>% 
  unnest(guest) %>% 
  anti_join(guest, by = c("guest" = "name")) %>% 
  knitr::kable(caption = "Guests with emails who are not on the guestlist")

```


# 3.3 

Make a guestlist that includes everyone we have emails for (in addition to those on the original guestlist). Again, separate the list of guests in 'email' into their own row before joining tibbles. 

```{r E3.3}
# make a super guestlist by joining both tibbles
# superlist will have all the folks from 'guest' tibble, whether we have their email or not, as well as folks who we have email for who weren't included on the guestlist.
full_join(guest, email %>% 
            mutate(guest = strsplit(as.character(guest), ", ")) %>% 
            unnest(guest), 
          by = c("name" = "guest")) %>% 
  DT::datatable(caption = "Super Guestlist: all wedding guests")
```

