---
title: 'Assignment 2: Gapminder and ''dplyr'''
author: "Hannah McSorley"
date: "21/09/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Homework Assignment 2

The Objective of this assignment is the practice using ggplot2 and dplyr to explore and visualize the gapminder dataset.

```{r}
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tidyverse))  # included package 'dplyr'
suppressPackageStartupMessages(library(tsibble))
suppressPackageStartupMessages(library(knitr))  # tidy tables
suppressPackageStartupMessages(library(DT))  # tidy dataframe tables for HTML
```

## 'dplyr' function focus
Within the 'dplyr' library, important functions to employ are:

* _filter()_: a tidy way to subset _in-situ_
* _%>%_: Piping to simplify multiple actions on a dataset

#1.1 filter()
Subset the 'gapminder' data to three countries in the 1970's using the 'filter()' function.

```{r filter}

# filter for the 1970's for three countries
gapminder %>% 
  dplyr::filter(between(year, 1970, 1979),
                country == "Puerto Rico" | country == "Yemen, Rep." | country == "Zimbabwe")

# countries were selected randomly using the function 'sample_n()' 
# sample_n() selects n number of rows randomly
# gapminder %>% filter(between(year, 1970, 1979)) %>% sample_n(size = 3)

# I want to use this filtered dataset again, so save it as a new object
seventies_three <- gapminder %>% 
  dplyr::filter(between(year, 1970, 1979),
                country == "Puerto Rico" | country == "Yemen, Rep." | country == "Zimbabwe")
```

#1.2 pipe
Use the pipe operator %>% to select “country” and “gdpPercap” from your filtered dataset in 1.1.

```{r pipe}

# subset the dataset created in 1.1
seventies_three %>% 
  select(country, gdpPercap)
```
# 1.3 Advanced Filtering 

Filter gapminder to all entries that have experienced a drop in life expectancy. Be sure to include a new variable that’s the increase in life expectancy in your tibble. Hint: you might find the lag() or diff() functions useful.

```{r lagfilter}
# calculate the difference in life expectancy over time for each country
# filter for those with a drop in life expectancy 
gapminder %>% 
  arrange(country, year) %>% 
  mutate(change_lifeExp = tsibble::difference(lifeExp)) %>% 
  filter(change_lifeExp < 0) %>% 
  head() %>%
  kable(digits = 2)
  
# unclear intruction about including a new variable for the increase in life expectancy... 
# when we're filtering for the drop in life expectancy

# maybe something that identifes whether there was an increase or decrease?
gapminder %>% 
  arrange(country, year) %>% 
  drop_na() %>% 
  mutate(change_lifeExp = tsibble::difference(lifeExp),
         incr_lifeExp = (change_lifeExp > 0)) %>%
  head() %>%  
  kable(digits = 2)
# provides TRUE or FALSE indicator of whether the lifeExp increased for that entry,
# relative to the previous data

# filter the above to include only entries that have had a drop in life expectancy
gapminder %>% 
  arrange(country, year) %>% 
  drop_na() %>% 
  mutate(change_lifeExp = tsibble::difference(lifeExp),
         incr_lifeExp = (change_lifeExp > 0)) %>% 
  filter(incr_lifeExp == FALSE) %>% 
  select(country:change_lifeExp) %>% 
  kable(digits = 2, caption = "drop in life expectancy")
```

# 1.4 Advanced Filtering

Filter gapminder so that it shows the max GDP per capita experienced by each country. Hint: you might find the max() function useful here.
```{r}
# create a variable for the maximum GDP per capita in each country
gapminder %>% 
  group_by(country) %>%
  mutate(max_gdp = max(gdpPercap)) %>% 
  filter(gdpPercap == max_gdp) %>% 
  head() %>% 
  kable(digits = 2)
```
## 1.4 Alternate
__OR__ 

Filter gapminder to contain six rows: the rows with the three largest GDP per capita, and the rows with the three smallest GDP per capita. Be sure to not create any intermediate objects when doing this (with, for example, the assignment operator). Hint: you might find the sort() function useful, or perhaps even the dplyr::slice() function.
```{r}
# use above code to add max gdpPercap
# filter to 3 highest max_gdp and 3 lowest max_gdp

# NOPE -- didn't get it this time.
# gapminder %>% 
  #group_by(country) %>%
  #mutate(max_gdp = max(gdpPercap)) %>% 
  #filter(gdpPercap == max_gdp) %>%
  #arrange(max_gdp) %>% 
  #filter(between(row_number(), 1, 3) & between(row_number(), n(), (n()-3)))

```


# 1.5

Produce a scatterplot of Canada’s life expectancy vs. GDP per capita using ggplot2, without defining a new variable. That is, after filtering the gapminder data set, pipe it directly into the ggplot() function. Ensure GDP per capita is on a log scale.
```{r}
# life expectancy vs. GDP per capita
# do not create a new variable, pipe filters directly into plot
gapminder %>% 
  filter(country == "Canada") %>% 
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point(colour = "royalblue1", shape = 16) +
  scale_x_log10("GDP per capita", labels = scales::dollar_format()) +
  ylab("Life Expectancy (yrs)") +
  theme_bw()
  
```


# Exercise 2

Explore individual variables in _dplyr_
Pick one __categorical variable__ and one __quantitative variable__ to explore.
Answer the following questions in whichever way you think is appropriate, using dplyr:

-1. What are possible values (or range, whichever is appropriate) of each variable?
-2. What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.
* Feel free to use summary stats, tables, figures.

## 1 -- categorical variable: continent (and countries in each)

```{r exercise2}
## What are possible values (or range, whichever is appropriate) of each variable
# --- 
# how many continents are there in the gapminder dataset?
str(gapminder$continent)
# what are they?
unique(gapminder$continent)

# how many countries are in each continent 
gapminder %>% 
  group_by(continent) %>% 
  summarize(countries = n()) %>% 
  kable(caption = "number of countries by continent, 'gapminder'",
        col.names = c("Continent", "Country Count"))

# plot continents with their GDP per capita and life expectancy 
# x axis on a log scale
# get fancy with facet wrapping (multiple panels)
gapminder %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) + 
  geom_point() + 
  facet_wrap(~continent) + 
  geom_point() + 
  geom_smooth(method = "loess") + 
  facet_wrap(~continent, ncol=1) +
  scale_x_log10("GDP per capita", labels = scales::dollar_format()) 
```
There are five continents in the gapminder dataset. Africa has the most countries, followed by America (North, Central, South), then Asia and Europe, and Oceania 
 has the fewest countries. 
 
 ## 2 -- quantitative variable: GDP
```{r}
# What values are typical? What’s the spread? What’s the distribution? Etc., 
# tailored to the variable at hand.

# check the 'summary' stats
gapminder %>% 
  group_by(country) %>% 
  summary() 

# long way
# by continent,
# what's the mean, relative standard deviation, mode, min and max GDP per capita?
gapminder %>% 
  group_by(continent) %>% 
  summarize(mean_gdpp = mean(gdpPercap),
            sd_gdpp = sd(gdpPercap),
            rsd_gdpp = (sd(gdpPercap) / mean_gdpp) * 100, 
            med_gdpp = median(gdpPercap),
            min_gdpp = min(gdpPercap),
            max_gdpp = max(gdpPercap)) %>% 
  kable(digits = 2, caption = "GDP per capita by continent")

# what if you first multiple by population to get GDP? 
gapminder %>%
  mutate(gdpBill = round(gdpPercap*pop/10^9, digits = 2)) %>% 
  group_by(continent) %>% 
  summarize(mean_gdp = mean(gdpBill),
            sd_gdp = sd(gdpBill),
            rsd_gdp = (sd(gdpBill) / mean_gdp) * 100, 
            med_gdp = median(gdpBill),
            min_gdp = min(gdpBill),
            max_gdp = max(gdpBill)) %>% 
  kable(digits = 2, caption = "GDP by continent")

# how about summarise GDP per capita and population?
gapminder %>% 
  group_by(continent) %>% 
  select(gdpPercap, pop, continent) %>% 
  summarise(mean_gdp = mean(gdpPercap),
          sd_gdp = sd(gdpPercap),
          rsd_gdp = (sd(gdpPercap) / mean_gdp) * 100, 
          mean_pop = mean(pop),
          sd_pop = sd(pop),
          rsd_pop = (sd(pop) / mean_pop) * 100) %>% 
  kable(digits = 2, caption = "GDP per capita and population")
# yuck. that's a lot of numbers


# try plotting it
# put GDP on a log scale again
gapminder %>% 
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  labs(y = "GDP per capita",
       x = "") +
  scale_y_log10("GDP per capita", labels = scales::dollar_format())

# plot population also, because I think they're related
gapminder %>% 
  ggplot(aes(x = continent, y = pop)) +
  geom_boxplot() +
  scale_y_log10("Population (log)")
```


# Exercise 3: Explore various plot types

Make two plots that have some value to them. That is, plots that someone might actually consider making for an analysis. Just don’t make the same plots we made in class – feel free to use a data set from the datasets R package if you wish.

* A scatterplot of two quantitative variables.
* One other plot besides a scatterplot.

Note: You don’t have to use all the data in every plot! It’s fine to filter down to one country or a small handful of countries.
```{r Exercise3}

# 2 scatter plots
# these are similar facet plots like I plotted in Exercise 2 -- because I like it

# life expectancy per capita over time
gapminder %>%
  ggplot(aes(x = year, y = lifeExp)) + 
  geom_point() +
  geom_smooth(method = "loess") +
  facet_wrap(~ continent, ncol = 1) +
  labs(x = "", y = "Life Expectancy")

# GDP per capita over time
gapminder %>%
  ggplot(aes(x = year, y = gdpPercap)) + 
  geom_point() + 
  geom_smooth(method = "loess") +
  facet_wrap(~ continent, ncol = 1) +
  scale_y_log10("GDP per capita", labels = scales::dollar_format()) +
  labs(x = "")

```

One other plot (not a scatter plot)

```{r histogram}

# histograms show distribution of a quantitative variable
# 'geom_freqpoly()'
# shows distribution like a histogram but with lines instead of shaded bars
gapminder %>% 
  ggplot(aes(x = lifeExp, color = continent)) +
  theme_bw() +
  labs(x = "Life Expectancy (yrs)", y = "distribution frequency count") +
  geom_freqpoly()

```
  
# Recycling (Optional) 

Rubrics:

    reasoning
        How well did you reason through the scenario?

For people who want to take things further.

Evaluate this code and describe the result. Presumably the analyst’s intent was to get the data for Rwanda and Afghanistan. Did they succeed? Why or why not? If not, what is the correct way to do this?

```{r}

# provided code
filter(gapminder, country == c("Rwanda", "Afghanistan")) %>% 
  kable(digits = 2, caption = "original code")  
# I added the kable call for easier comparison between this and my code below
```
It does not appear that this method of filtering delivered the desired results. It looks like data from Rwanda and Afghanistan for overlapping years were ommitted, resulting in each country having data provided every decade: Afghanistan had data for every ten years ending in 7 (e.g. 1957, 1967, 1977, etc.) while Rwanda had data for every ten years ending in 2 (e.g. 1952, 1962, 1972, etc.).
I think the problem is associated with trying to filter country based on a vector of names. When the country filter is explicitly defined as Rwanda OR Afghanistan, the result is a tibble twice as long as the original output. This method results in data for both countries on overlapping years.

```{r}
# try some other options to see if the results agree
gapminder %>% 
  filter(country == "Rwanda" | country == "Afghanistan") %>% 
  kable(digits = 2, caption = "modified code")
``` 


# Tibble display (Optional) 

Present numerical tables in a more attractive form using knitr::kable() for small tibbles (say, up to 10 rows), and DT::datatable() for larger tibbles.

_I believe I incorporated this into my code already_ 
:smiley:


