---
title: "STAT 545 hw-01 (gapminder)"
author: "Hannah J McSorley"
date: "September 11, 2019"
output: ioslides_presentation  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Homework 1, Exercise 3

This is the third exercise in assignment one for STAT 545. The goal of this exercise is to create slides using R Markdown. 

## 'gapminder' dataset 

This exercise uses the same analysis of the 'gapminder' dataset as in hw01 Exercise 2. 

```{r load_packages, warning = FALSE}
# load the packages required for this analysis
library(gapminder)
```

## Data structure 

Explore the data structure and type by executing some simple commands.

```{r gapminder}
str(gapminder)  # structure
```

Calling 'str()' yields lots of information about this dataset (length, data structure and type).

## Dataset Exploration: Countries
 
Identify how many countries are included in the gapminder dataset, and what they are.
```{r countries}
# see if Canada is a country in the dataset
any(gapminder$country == "Canada")

#  how many countries are included?
length(unique(gapminder$country))
```
_Note:_ the number of countries was also indicated in the 'str(gapminder)' call.

## Dataset Exploration: Span of Dataset (Years)

This dataset includes life expectancy ("lifeExp") in each country. What is the range of this data and how does it vary? Using the 'range()' function yields the lowest and highest end members in the data (min & max).

```{r years}
# what is the range of years included?
range(gapminder$year)

# what is the range of life expectancies across the world?
range(gapminder$lifeExp)
```

## Dataset Exploration: Life Expectancy

Identify countries with the lowest and highest life expectancy, and associated information.

```{r lifeExp}
# subset using the index command 
# minimum life expectancy
gapminder[which(gapminder$lifeExp == min(gapminder$lifeExp)), ]
```
The lowest life expectancy (23.6 years) was in 1992 in Rwanda, Africa. 

## Dataset Exploration: Life Expectancy

```{r}
# maximum life expectancy?
gapminder[which(gapminder$lifeExp == max(gapminder$lifeExp)), ]
```

The highest life expectancy (82.6 years) was in Japan, Asia in 2007. 


## Dataset Exploration: GDP & Life Expectancy

### Is there a positive relationship between life expectancy and GDP? Is the relationship linear?


To visually explore the relationship between GDP and life expectancy, plot those data against each other. 


##
```{r plot}
# use base R to plot global life expectancy with GDP
plot(y = gapminder$lifeExp, x = gapminder$gdpPercap,
     ylab = "life expectancy (yrs)", xlab = "GDP percapita")
```

## Dataset Exploration: Canada, eh 

When was Canadian life expectancy lowest and highest?

```{r eh}
# subset the gapminder data for Canada
Canada_df <- gapminder[gapminder$country == "Canada", ]

# find minimum and maximum life expectancy
Canada_df[which(Canada_df$lifeExp == min(Canada_df$lifeExp)), ]
```
The life expectancy in Canada was lowest in the first recorded year of this dataset (1952, 68.8 years).

## Canada, eh
```{r}
Canada_df[which(Canada_df$lifeExp == max(Canada_df$lifeExp)), ]
```

Canadian life expectancy was highest in the most recent recorded year (2007, 80.7 years). 

## 
Life expectancy in Canada has steadily increased over time from 1952-2007. 
```{r plot_eh}
# use base R to plot Canadian life expectancy over time
plot(y = Canada_df$lifeExp, x = Canada_df$year,
     ylab = "life expectancy (yrs)", xlab = "Year",
     type = "b")
```

